ECCCos from the Black Box: Letting Models speak for Themselves

### 2023-04-28 notebooks/mnist.qmd

plts = []
for (_name,ce) in ces
    _x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plts = plts[plt_order]
plts = [p1, plts...]
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "mnist_eccco_benchmark.png"))
```

@@

>>>>

plts = []
for (_name,ce) in ces
    _x = CounterfactualExplanations.counterfactual(ce)
    _phat = target_probs(ce)
    _title = "$_name (p̂=$(round(_phat[1]; digits=3)))"
    plt = Plots.plot(
        convert2image(MNIST, reshape(_x,28,28)),
        axis=nothing, 
        size=(img_height, img_height),
        title=_title
    )
    plts = [plts..., plt]
end
plts = plts[plt_order]
plts = [p1, plts...]
plt = Plots.plot(plts...; size=(img_height*length(plts),img_height), layout=(1,length(plts)))
display(plt)
savefig(plt, joinpath(output_images_path, "mnist_all_generators.png"))
```

@@

```{julia}
df = @chain bmk() begin
    @filter(variable in [
        "distance_from_energy",
        "distance_from_targets",
        "distance",])
    @mutate(variable = ifelse.(variable .== "distance_from_energy", "Non-Conformity", variable))
    @mutate(variable = ifelse.(variable .== "distance_from_targets", "Implausibility", variable))
    @mutate(variable = ifelse.(variable .== "distance", "Cost", variable))
end
plt = AlgebraOfGraphics.data(df) * visual(BoxPlot) * 
    mapping(:generator, :value, row=:variable, col=:model, color=:generator)
plt = draw(
    plt, axis=(xlabel="", xticksvisible=false, xticklabelsvisible=false, width=150, height=120), 
    facet=(; linkyaxes=:minimal)
)   
display(plt)
save(joinpath(output_images_path, "mnist_benchmark.png"), plt, px_per_unit=5)
```

### 2023-04-28 paper/paper.tex

\begin{minipage}[c]{0.40\textwidth}
  \centering
  \includegraphics[width=\textwidth]{../artifacts/results/images/eccco_illustration.png}
  \captionof{figure}{Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.50\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}

@@

\begin{minipage}[c]{\textwidth}
  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
  \captionof{figure}{Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
\end{minipage}

@@

The entire procedure for Generating ECCCos is described in Algorithm~\ref{alg:eccco}. For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state. That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}. Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.

@@

\section{Experiments}\label{conformity}

@@

\section{Experiments}

@@

>>>>

\begin{minipage}[c]{0.40\textwidth}
  \centering
  \includegraphics[width=\textwidth]{../artifacts/results/images/eccco_illustration.png}
  \captionof{figure}{[PLACEHOLDER] Vector fields indicating the direction of gradients with respect to the different components of the ECCCo objective (Equation~\ref{eq:eccco}).} \label{fig:eccco}
\end{minipage}  
\hfill
\begin{minipage}[c]{0.50\textwidth}
\captionof{algorithm}{Generating ECCCos (For more details, see Appendix~\ref{app:eccco})}\label{alg:eccco}
\begin{algorithmic}[1]
  \Require $\mathbf{x}, \mathbf{y}^*, M_{\theta}, f, \Lambda, \alpha, \mathcal{D}, T, \eta, n_{\mathcal{B}}, N_{\mathcal{B}}$ \linebreak where $M_{\theta}(\mathbf{x})\neq\mathbf{y}^*$
  \Ensure $\mathbf{x}^\prime$
  \State Initialize $\mathbf{z}^\prime \gets f^{-1}(\mathbf{x})$
  \State Generate buffer $\mathcal{B}$ of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld})
  \State Run \textit{SCP} for $M_{\theta}$ using $\mathcal{D}$
  \State Initialize $t \gets 0$
  \While{\textit{not converged} or $t < T$}
  \State $\hat{\mathbf{x}}_{\theta, t} \gets \text{rand}(\mathcal{B},n_{\mathcal{B}})$
  \State $\mathbf{z}^\prime \gets \mathbf{z}^\prime - \eta \nabla_{\mathbf{z}^\prime} \mathcal{L}(\mathbf{z}^\prime,\mathbf{y}^*,\hat{\mathbf{x}}_{\theta, t})$
  \State $t \gets t+1$
  \EndWhile
  \State $\mathbf{x}^\prime \gets f(\mathbf{z}^\prime)$
\end{algorithmic}
\end{minipage}

@@

\begin{minipage}[c]{\textwidth}
  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_eccco.png}
  \captionof{figure}{[SUBJECTO TO CHANGE] Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:eccco-mnist}
\end{minipage}

@@

The entire procedure for generating ECCCos is described in Algorithm~\ref{alg:eccco}. For the sake of simplicity and without loss of generality, we limit our attention to generating a single counterfactual $\mathbf{x}^\prime=f(\mathbf{z}^\prime)$ where in contrast to Equation~\ref{eq:eccco} $\mathbf{z}^\prime$ denotes a $1$-dimensional array containing a single counterfactual state. That state is initialized by passing the factual $\mathbf{x}$ through the encoder $f^{-1}$ which in our case corresponds to a simple feature transformer, rather than the encoder part of VAE as in REVISE~\citep{joshi2019realistic}. Next, we generate a buffer of $N_{\mathcal{B}}$ conditional samples $\hat{\mathbf{x}}_{\theta}|\mathbf{y}^*$ using SGLD (Equation~\ref{eq:sgld}) and conformalise the model $M_{\theta}$ through Split Conformal Prediction on training data $\mathcal{D}$.

@@

\section{Empirical Analysis}\label{emp}

@@

\subsection{Data}

\subsection{Results}

\begin{figure}
  \includegraphics[width=\textwidth]{../artifacts/results/images/mnist_benchmark.png}
  \caption{[SUBJECTO TO CHANGE] Original image (left) and ECCCos for turning an 8 (eight) into a 3 (three) for different Black Boxes from left to right: Multi-Layer Perceptron (MLP), Ensemble of MLPs, Joint Energy Model (JEM), Ensemble of JEMs.}\label{fig:mnist-benchmark}
\end{figure}

\subsection{Key Insights}

\subsection{Limitations}

@@

\section{Conclusion}
